COMP 6721: Project Report

### 2019-03-25 report/paper.tex
Add part for vspace.

\begin{document}
%
\title{COMP 6721: Project Report}
\author{Martin Khannouz}
%
\institute{40093517 \email{martin.khannouz@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\section*{Introduction}
This project report describes how the first
project of \textit{Introduction to Artificial
Intelligence} were developed. The first section
covers the environment and the tools used for the
project. The heuristic section describes each
heuristic and how they were improved. The
difficulties section summerizes the difficulties
encountered during the project and how they were
overcome.

\section{Settings and Structures}
\subsection{Settings}
The developing part of this project was realized
on an Archlinux distribution with Python~3.7.2.
Vim with a Jedi plugin was used for the coding
part and a private Github repository was set to
keep track of the changes.
The \textit{orwell} machine was used before each
deliverable to ensure the project worked with the
software version on Concordia's computers. Pytest
was used to guarantee no regression has been made.

\subsection{stucture}
The project is structured in classes with
the main one being the engine. This class is an
instance of the game that implements the rules.
To ensure this class operates properly, unit tests
were implemented before the class methods, based on
the use cases given by the teacher.
This class was optimized as much as possible to
guarantee the fastest execution time for minimax
player.

@@

In addition to these classes, additional functions
were implemented. One function used genetic
algorithms to optimize the \textit{Vspace}
heuristic, but it was too slow and just gave hint
on how to enhance the heuristic rather than a
accurate weights.  A win rate function was also
implemented to compare combinations of player and
heuristic against each other.

@@

\subsection{Convolution}
The convolution utilize the \textit{convolve2d} function
from Scipy to estimate the value of a current
state.
The game board is split into two matrices: dot and
color. Each matrix contains -1, 0, or 1. A zero
indicates an empty space while the two other values
represent either a filled dot and an empty dot, or
a white cell and a red cell.
Applying \textit{convolve2d} on these matrices
produces new matrices with values that range from
-4 to 4. An absolute value of four indicates a row
of four and therefore, a winning position.
The heuristic gathers all intermediate values into
an histogram (using the \textit{histogram}
function of numpy) then give a weight to each
value.

Figure~\ref{fig:board-example} shows an example of
the board. Figure~\ref{fig:dot-representation}
focus on the dot player and show the
transformation applied to the board for this
player. Filled dots are set to 1 and circles to
-1. Zeros are not represented for more clarity.
The heuristic applies four convolutions: one for
the rows, one for the columns and one for each
diagonal. These convolution generate four smaller
matrices showed in
Figure~\ref{fig:convolution-results}.  Then, the
matrices are flatened, concateneted and the
absolute value is applied to every cell. Finally
the histogram is built out of this matrix and each
element is multiplied by a weight before being
summed.  The final
value of the board is the difference between the
two player values.

This approach is strong for many reason. It is
faster than calling python code because Scipy and
Numpy rely on C writen functions. Uncomplete rows
that contain zeros are more valued than uncomplete
rows that contain opposite values. 
\begin{figure}[ht]
		\center
		\begin{tikzpicture}

@@

\subsection{Vspace}
The Vspace heuristic tries to give value to empty
cells in addition to counting rows of similar
symbols. This heuristic count the number of lines
and their sizes.
\section{Difficulties}
\begin{itemize}
		\item Slowness of python, (numpy first)
		\item Version changing between Python 3.7.2 and 3.5.1 (serialization of regex)
\end{itemize}
\section{Analyze}
\begin{itemize}
		\item How is the pruning.
		\item How are the performance at the tournament.
\end{itemize}
\section{A Subsection Sample}
Please note that the first paragraph of a section or subsection is
not indented. The first paragraph that follows a table, figure,
equation etc. does not need an indent, either.

>>>>

\begin{document}
%
\title{COMP 6721: Project Report}
\author{Martin Khannouz}
%
\institute{40093517 \email{martin.khannouz@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\section*{Introduction}
This project report describes how the first
project of \textit{Introduction to Artificial
Intelligence} was developed. The first section
covers the environment and the tools used for the
project. The heuristic section describes each
heuristic and how they were improved. The
difficulties section summerizes the difficulties
encountered during the project and how they were
overcome.

\section{Settings and Structures}
\subsection{Settings}
The developing part of this project was realized
on an Archlinux distribution with Python~3.7.2.
Vim with the Jedi plugin was used to code and a private Github repository was set to
keep track of the changes.
The \textit{orwell} machine was used before each
deliverable to ensure the project worked with the
software version on Concordia's computers. Pytest
was used to guarantee no regression has been made.

\subsection{stucture}
The project is structured in classes with
the main one representating the engine. This class is an
instance of the game and implements the rules.
To ensure this class operates properly, unit tests
were implemented before the class methods, based on
the use cases given by the teacher.
This class was optimized as much as possible to
guarantee the fastest execution time for minimax
algorithm.

@@

In addition to these classes, additional functions
were implemented. One function used genetic
algorithms to optimize the \textit{Vspace}
heuristic, but it was too slow and just gave hint
on how to enhance the heuristic rather than a
accurate weights.  A win rate function was also
implemented to compare combinations of player and
heuristic against each other. Finally, a non
tweaked MinMax version was developed for
deliverable 2.

@@

\subsection{Convolution}
The convolution utilize the \textit{convolve2d} function
from Scipy to estimate the value of a current
state.
The game board is split into two matrices: dot and
color. Each matrix contains -1, 0, or 1. A zero
indicates an empty space while the two other values
represent either a filled dot and an empty dot, or
a white cell and a red cell.
Applying \textit{convolve2d} on these matrices
produces new matrices with values that range from
-4 to 4. A cell with an absolute value of four indicates a row
of four and therefore, a winning position.
The heuristic gathers all intermediate values into
an histogram (using the \textit{histogram}
function of numpy) then give a weight to each
value. For instance, a weight of 16 was given to
each intermediate value of 3 which represent a row
of three.

Figure~\ref{fig:board-example} shows an example of
the board. Figure~\ref{fig:dot-representation}
focus on the dot player and show the
transformation applied to the board for this
player. Filled dots are set to 1 and circles to
-1. Zeros are not represented for more clarity.
The heuristic applies four convolutions: one for
the rows, another for the columns and one for each
diagonal. These convolution generate four smaller
matrices showed in
Figure~\ref{fig:convolution-results} (left side).
The matrices are then flatened, concateneted and the
absolute value is applied to every cell. Finally
the histogram is built out of this matrix and each
element is multiplied by a weight before being
summed.  The final
value of the board is the difference between the
two player values.

This approach is strong for many reason. It is
faster than calling python code because Scipy and
Numpy rely on C writen functions. Uncomplete rows
that contain zeros are more valued than uncomplete
rows that contain opposite values. Finally,
because the convolution mask is applied to empty
spaces, the heuristic can build complexe
structures with space in the middle or above.

\begin{figure}[ht]
		\center
		\begin{tikzpicture}

@@

\clearpage
\subsection{Vspace}
The goal of the Vspace heuristic is to give value to empty
cells in addition to counting rows of similar
symbols. Similarly to the previous heuristic,
Vspace count the size of the lines, but for every
line connected to an empty cell, it uses the
size of this line as a value for the empty cell.

Figure~\ref{fig:dot-vspace} and
Figure~\ref{fig:color-vspace} give examples on
how the spaces are valued for both players, based
on the board at Figure~\ref{fig:board-example}.
Figure~\ref{fig:color-vspace} shows that even if
the line is split in two with an empty cell in the
middle, the combine value of both side gives the
value of the space. This is the reason why the two
3-values appears even though there is no line of
three.

Possible value for spaces range from 0 to 4 and
higher values are downed to 4. When multiple lines
are connected to an empty cell, the cell receives
the maximum value from them. The heuristic split the
space in two categories: the space available in
the next turn and the spaces too high to be reach
with one card.

Finally, an histogram for each category is made and
each value is multiplied by a weight then summed.
Note that the two categories have different
weights, because a space connected to a line of
three symboles does not hold the same value if you
can play on it on the next turn or not.
\begin{figure}[ht]
		\center
		\begin{tikzpicture}

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0, 0) {$\bullet$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 0) {$\circ$};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 0) {};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 0) {$\bullet$};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (2.0, 0) {};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0, 0.5) {$\circ$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 0.5) {$\bullet$};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 0.5) {};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 0.5) {$\circ$};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (2.0, 0.5) {};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0, 1) {$\circ$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 1) {$\circ$};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 1) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 1) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] (A) at (2.0, 1) {};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0, 1.5) {$\bullet$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 1.5) {$\bullet$};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 1.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 1.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (2.0, 1.5) {};

				%------
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5, 0) {$\bullet$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 0) {$\circ$};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 0) {1};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 0) {$\bullet$};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 0) {1};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5, 0.5) {$\circ$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 0.5) {$\bullet$};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 0.5) {1};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 0.5) {$\circ$};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 0.5) {1};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] (B) at (5, 1) {$\circ$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 1) {$\circ$};
				\node [draw, fill=purple!25, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 1) {2};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 1) {1};
				\node [draw, fill=purple!25, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 1) {1};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5, 1.5) {$\bullet$};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 1.5) {$\bullet$};
				\node [draw, fill=purple!25, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 1.5) {2};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 1.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 1.5) {};

				\draw [->,thick] (A) -- (B);
		\end{tikzpicture}
		\caption{Example of the Vspace process for the
		dot player based on the board in
		Figure~\ref{fig:board-example}. the grey cells
		represent cells that contain nothing
		interesting. On the right board, green cells
		are spaces available for the next turn and
		purple cells are spaces that need at least two
		cards to be filled.}
		\label{fig:dot-vspace}
\end{figure}
\begin{figure}[ht]
		\center
		\begin{tikzpicture}

				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (0, 0) {};
				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 0) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 0) {};
				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 0) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (2.0, 0) {};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0, 0.5) {};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 0.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 0.5) {};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 0.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (2.0, 0.5) {};

				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (0, 1) {};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 1) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 1) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 1) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (2.0, 1) {};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (0, 1.5) {};
				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (0.5, 1.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.0, 1.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (1.5, 1.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (2.0, 1.5) {};

				%------
				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (5, 0) {};
				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 0) {};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 0) {3};
				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 0) {};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 0) {1};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5, 0.5) {};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 0.5) {};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 0.5) {3};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 0.5) {};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 0.5) {1};

				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] (B) at (5, 1) {};
				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 1) {};
				\node [draw, fill=purple!25, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 1) {1};
				\node [draw, fill=green!15, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 1) {1};
				\node [draw, fill=purple!25, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 1) {1};

				\node [draw, minimum width=0.5cm, minimum height=0.5cm] at (5, 1.5) {};
				\node [draw, fill=red!20, minimum width=0.5cm, minimum height=0.5cm] at (5.5, 1.5) {$\bullet$};
				\node [draw, fill=purple!25, minimum width=0.5cm, minimum height=0.5cm] at (6.0, 1.5) {2};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (6.5, 1.5) {};
				\node [draw, fill=black!20, minimum width=0.5cm, minimum height=0.5cm] at (7.0, 1.5) {};

				\draw [->,thick] (A) -- (B);
		\end{tikzpicture}
		\caption{Example of the Vspace process for the
		color player based on the board in
		Figure~\ref{fig:board-example}. the grey cells
		represent cells that contain nothing
		interesting. On the right board, green cells
		are spaces available for the next turn and
		purple cells are spaces that need at least two
		cards to be filled.}
		\label{fig:color-vspace}
\end{figure}
\clearpage
\section{Difficulties}
\begin{itemize}
		\item Slowness of python, (numpy first)
		\item Version changing between Python 3.7.2 and 3.5.1 (serialization of regex)
\end{itemize}
\section{Analyze}
\begin{itemize}
		\item How is the pruning depth 3 with a
				recycling moves: no prun = 26s, prun=0.3s.
		\item How are the performance at the tournament.
\end{itemize}

\section{A Subsection Sample}
Please note that the first paragraph of a section or subsection is
not indented. The first paragraph that follows a table, figure,
equation etc. does not need an indent, either.
